Open Questions for Computer Science and Cheminformatics

### 2011-11-17 OpenQuestions.tex
Minor rewrite

The key bonus of the semantic web is that different data sources can be readily integrated with each other. In the field
of Cheminformatics, we are not only interest in linking two molecules 
(the normalization problem for different protomers, tautomers, or special cases of isomerisms remains open), but we
are also interested in being able to search efficiently within molecules when being linked via semantic web approaches.
What could be algorithmic solutions for this?
\end{enumerate}

@@

\section*{Data mining}
\begin{itemize}
\item evaluation of similarities in a heterogeneous network.  What is a specific example here?
\item integrate chemical structure information with ontologies.  Specific example problem?
\item Classify molecular descriptors up to equivalence, and dependence on one another.
\item systems-level understanding of small molecules.  What does this mean, and what would a specific challenge problem be?
\end{itemize}

\begin{enumerate}

\item \emph{Efficient molecule browsing, e.g. on scaffold level}.

Chemical Abstract Services have a molecule browsing tool called SubScape, which allows to brows large-scale
chemical spaces efficiently. What could be large-scale solutions for doing this within (combined and aligned) 
public databases.
%
\item \emph{Large-scale browsing of molecular property spaces, e.g. on scaffold level, side-effect-level, ...}.

Certain molecules might have hundreds of biological activities, side-effects in humans (from clinical trials), or
many other properties attached to them. What are large-scale mining and visulization options, especially when thinking
about mining private and public datasources at the very same time?
%
\item \emph{Patent text mining (curation)}.

There are various tools for doing automatic text mining on chemical patents. Still, the overall acceptance rate is improvable
since many medicinal chemists are very concerned about the data quality of such efforts. What could be done to improve
the mining quality and to provide confidence level estimations for each molecule coming from patent mining? 
How can patent mining be used to create new drugs faster or to speed-up collaboration/licensing discussions?
\end{enumerate}

\section*{Machine learning}
I don't know.  Help?  Ideas: better kernelization techniques (something specific though), better algorithms to train X model about Y thing.
\begin{enumerate}
\item \emph{Large-scale vectorial versus kernels molecule similarity}

Vectorial molecule encodings serve as efficient approximations.
Sometimes non-vectorial molecular 3D shape or molecule kernel comparisons might be more suitabe to compare molecules, since
they might better correlate with biological activity, toxicity in humans, etc. One key problem is that molecular 3D shape 
or molecule kernel approaches require to compare two molecules (or their 3D conformational eplosions) directly.
This becomes prohibitively expensive when considering millions of molecules. What could be potential solutions for approximating boundary 
conditions or large-scale mining methods allowing within a second range to return all similar molecules to molecule X, especially
considering not just one vectorial encoding, but molecule kernels, ligand-protein similarities (for example from ligand-protein crystal structures), 
or chemogenomics similarities (all reported biological activities for one molecule).
%
\item \emph{Using multiple annotations for improving molecular mining/predictions (chemogenomics)}

As an example: Biological activities might not be independent of each other, but have a certain correlation between each other.
In Chemogenomics this is used for creating models of combining molecules with protein sequences, molecules with active sites of proteins,
molecules with biological read-outs of multiple assays. How can we optimize such highly complex mining scenarios, especially
when considering large-scale data sets with hundred of thousands molecules and thousands of biological activities? 
How can we combine, mine, and visualize categorial and continuous output variables, e.g. hydrophobicity of a molecule and toxicity in humans, 
by still being able to make conrete proposals to medicinal chemistry of which parts of the molecule needs to be changed to
optimize the effect on a certain set of multi-objective variables? Is analoging (creating vrey small modifications of a molecule and measuring its activities)
really the most efficient way forward? If we test molecules, should we test it in a single biological assay or in multiple biological assays, if multiple, which ones?
If a company does not have a biological assay within reach, which other partner could offer testing a molecule within two days (vendor matching based on licenses)?
\end{enumerate}

@@

There exist a multitude of programs generation 3D conformations of molecules. Still, often input formats vary and require
to re-create 3D conformational spaces with other programs. This does not only complicate the maintenance scenario of such
databases, especially with the explosion of public databases, but also creates hugh space and time complexity issues when
searching within such databases. What could be better interfaces, maintenance, data structures, and private/public sharing
scenarios for conformational 3D databases? 
Many software vendors use different solutions for parallizing comput jobs: SGE, PVM, MPI, etc. 
Everyone knowing thh enterprise structure within companies might know that having more than one parallel processing framework 
is not easy? What could be better ways to streamline parallel processing structures for 
cheminformatics (and molecular modeling) algorithms? Is a cloud really an option? What about SaaS with secured data transfer?
\end{enumerate}

@@

Having within an organization, e.g. a commercial company, single or a small number of established KM and ELN products. 
How can we improve the maintenance, leveraging, and collaboration with many external partners (each of them potentially
with another KM/ELN solution)? Which party is hosting which data in which data structure, and how can we ensure that only
pre-defined data entries (and a limited number of annotations, e.g. bioological activities) are visible to a partner.
How can this be organized for a multitude of partners? Cloud computing?
\end{enumerate}

>>>>

One key concept of the linked data web, the semantic web, is that different data sources can be readily integrated with each other. Still, in the field
of Cheminformatics, we are not only interest in linking two molecules 
(the linking normalization problem for different protomers, tautomers, or special cases of isomerisms remain open), but we
are also interested in being able to search efficiently within molecules when being linked via semantic web approaches.
What could be algorithmic solutions for this?
\end{enumerate}

@@

\section*{Data mining}
\begin{enumerate}
\item evaluation of similarities in a heterogeneous network (JKW: What are heterogenous networks?).  What is a specific example here?
\item integrate chemical structure information with ontologies.  Specific example problem?
\item Classify molecular descriptors up to equivalence, and dependence on one another.
\item systems-level understanding of small molecules (JKW: which systems? This will help being clearer on the challenge).  What does this mean, and what would a specific challenge problem be?
%\item \emph{Efficient molecule browsing, e.g. on scaffold level}.
%
%Chemical Abstract Services have a molecule browsing tool called SubScape, which allows to browse large-scale
%chemical spaces efficiently. What could be large-scale solutions for doing this within (combined and aligned) 
%public databases.
%
\item \emph{Large-scale browsing of molecular property spaces, e.g. on scaffold level, side-effect-level, ...}.

Certain molecules might have hundreds of biological activities, side-effects in humans (SIDER database \cite{Kuhn_Campillos_Letunic_Jensen_Bork_2010}), or
many other properties attached to them. What are large-scale mining and visulization options? 
How can we mine private and public data sources at the very same time?
%
\item \emph{Patent text mining (curation)}.

There are various tools for doing automatic text mining on chemical patents. Still, the overall acceptance rate of chemical
text mining is improvable, since many medicinal chemists are very concerned about the data quality of such efforts. 
What could be done to improve the mining quality, curate the obtained data, and to provide confidence level estimations 
for each molecule coming from patent mining? Do require image2structure and text2structure mining also data stores
for ensuring a sufficient amount of confidence and data quality?
How can patent mining be used to create new drugs faster or to speed-up collaboration/licensing discussions?
\end{enumerate}

\section*{Machine learning}
\begin{enumerate}
\item \emph{Large-scale vectorial versus kernels molecule similarity}

Vectorial molecule encodings can serve as efficient approximations of molecules.
Sometimes non-vectorial molecular 3D shape or molecule kernel comparisons might be more suitabe to compare molecules, since
they might better correlate with activities. One key problem is that non-vectorial encodings require to compare all molecules
(or their 3D conformational eplosions) in a pair-wise manner.
This becomes prohibitively expensive when considering millions of molecules.
Can dyadic data approaches help \cite{Hochreiter:2006:SVM:1159508.1159516}? Other approximations or cascading flows?
%
\item \emph{Using multiple annotations for improving molecular mining/predictions (chemogenomics)}

As an example: Biological activities might not be independent of each other, but have a certain correlation between each other.
In Chemogenomics this is used for creating models of combining molecules with protein sequences, molecules with active sites of proteins,
or molecules with biological activities of multiple assays. How can we optimize such highly complex mining scenarios, especially
when considering large-scale data sets with hundred of thousands molecules and thousands of biological activities? 
How can we combine, mine, and visualize categorial and continuous output variables, e.g. hydrophobicity of a molecule and toxicity in humans, 
by still being able to make concrete proposals to medicinal chemistry? Is analoging (creating very small modifications of a molecule and measuring its activities)
really the most efficient way forward? If we test molecules, should we test it in a single biological assay or in multiple biological assays, if multiple, which ones?
If a company does not have a biological assay within reach, which other partner could offer testing a molecule within two days (vendor matching based on licenses or contracts)?
\end{enumerate}

@@

There exist a multitude of programs generation 3D conformations of molecules. Still, often input formats vary and require
to re-create 3D conformational spaces with other programs. This does not only complicate the maintenance scenario of such
databases, especially with the explosion of public databases, but also creates hugh space and time complexity issues when
searching within such databases. What could be better interfaces, maintenance, data structures, and private/public sharing
scenarios for conformational 3D databases? 
Many software vendors use different solutions for parallizing compute jobs: SGE, PVM, MPI, etc. 
Everyone knowing the enterprise IT approval cycles might know that having more than one parallel processing framework 
is not easy. What could be better ways to streamline parallel processing structures for 
cheminformatics (and molecular modeling) algorithms? Is a cloud really an option? What about SaaS with secured data transfer?
Can this also offer alternative licensing strategies for software suites in this domain?
\end{enumerate}

@@

Let us assume an organization, e.g. a commercial company, has a single or a small number of established KM and ELN products. 
How can we improve the maintenance, leveraging, and collaboration with many external partners (each of them potentially
with another KM/ELN solution)? Which party is hosting which data in which data structure (ontologies?), and 
how can we ensure that only pre-defined data entries 
(and a limited number of annotations, e.g. bioological activities) are visible to a partner.
How can this be organized for a multitude of partners? Cloud computing, user management, encryption granularity and efficient security management?
\end{enumerate}
